Nearest Neighbour Methods for Length of Stay Prediction

### 2014-12-23 papers/paper.tex
Rearrange paper structure, change Abstract to Summary

\documentclass[10pt]{article}

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage{natbib}
\usepackage{subcaption}

@@

\begin{abstract}
Predicting the length of hospital stay for injured patients involves making a
judgement about how long they need to be hospitalised, so that doctors and
hospital staff can plan the appropriate course of treatment for the patient
to ensure they recover in the quickest possible time. However, previous studies
have used a limited number of statistical techniques such as logistic
regression in order to predict the length of stay, and have not considered
automatic methods of feature selection.

Our work systematically evaluates and compares the performance of a number of
state-of-the-art classification algorithms and feature selection methods,
and also assesses the effect of feature discretisation for all combinations
of classifiers and feature selectors. We propose a new classification algorithm
called Ranked Distance Nearest Neighbor that considers the
relative importance of each feature when determining the nearest neighbours.

We were able to achieve improved accuracy and area
under the curve for both data sets, and we also show that our proposed Ranked
Distance Nearest Neighbour improves accuracy and area under the curve over
traditional nearest neighbour algorithms.
\end{abstract}

@@

\section{Problem Statement}
Our problem is a binary \textit{classification} task in which we
predict one of two possible categories for a patient: a length of stay (LOS) of
less than or equal to 2 days, or a LOS of greater than 2 days.

\section{Ranked Distance Nearest Neighbour}
\subsection{Intuition}
As we mentioned earlier, all features are considered equally in a $k$-NN
classifier, which implies that all features are equally important in predicting
the class. In practice, however, this is not always the case: to predict the
LOS of a trauma patient, the severity of their injury should affect the final
LOS more than whether or not they can speak English (which could be just two
of many features that are recorded about a patient). Our contribution is a
new NN algorithm that uses a modified distance function which takes into
account the relative importance of all features called Ranked Distance
Nearest Neighbour (RD).

\subsection{Mathematical formulation}
Given two instances $\mathbf{x} = (x_1,x_2,\ldots,x_n)$ and
$\mathbf{y} = (y_1,y_2,\ldots,y_n)$, the distance between them is given by:
\begin{equation*}
\mathrm{Distance} = \sum_{i=1}^n w_i |x_i-y_i|
\end{equation*}
where the $w_i$s weight the contribution of the $i$-th feature to the overall
distance that is computed by the $k$-NN algorithm. We assume that all feature
values have been normalised, as described above.

\subsection{Assignment of weights}
The weights $w_i$ can be tuned to match the particular problem that is being
investigated. For our LOS classification problem, we will consider one way
of weighting features.
First, we compute the correlation coefficients as described in the point above.
This gives us a ranking of the importance of each feature to the prediction of
the class, with 1 being the highest rank (indicating the most importance) and
$n$ being the lowest rank (indicating the least importance).
Instead of using the values of the correlation directly to weight
the distance contribution of each feature, we specify a function:
\begin{equation*}
f : \mathrm{Rank} \rightarrow \mathbb{R}, \mathrm{Rank} \in \{1,2,\ldots,n\}
\end{equation*}
that describes
how the weights of the features vary with the rank. This function should
inituitively be non-increasing and should produce lower values when the rank
number is greater, meaning that less important features have a lower weight
in the distance calculation.
Note that choosing $f(\mathrm{Rank}) = k$ for any non-zero constant $k$
results in all features contributing equally to the distance calculation.
We will use $f(\mathrm{Rank}) = \frac{1}{\mathrm{Rank}^c}$, where
$c$ varies from 0 to 1.

\section{Feature Selection}
Here we introduce the feature selection algorithms we evaluate in our work.
Several of these have not been investigated in previous work on length of
stay prediction.
\subsection{Correlation-based Feature Selection}
Correlation-based feature selection (CFS) is a method developed by Hall
\cite{Hall2000} which looks for a subset of features that have high correlation
with the class but low intercorrelation among the features. In this case,
correlation between two features is measured using what is called the
\textit{symmetric uncertainty}:
\begin{equation*}
U(A,B) = 2\dfrac{H(A)+H(B)-H(A,B)}{H(A)+H(B)}
\end{equation*}
where $A$ and $B$ are two nominal attributes and $H$ is the entropy function.
The entropies are calculated from the probabilities of each feature value, and
$H(A,B)$, the joint entropy of $A$ and $B$, is calculated from the
probabilities of each combination of values of $A$ and $B$. Note that $U(A,B)$
is always between 0 and 1.
CFS conducts a search through subsets of all features in order to find the
one feature subset that has the best merit.

\subsection{Pearson Correlation Coefficient}
The Pearson correlation coefficient measures the degree of linear correlation
between two variables. It is always in the range $[-1,1]$, where -1 is total
negative correlation, 0 is no (linear) correlation, and 1 is total positive
correlation. We compute the Pearson correlation between each feature and the
class using:
\begin{equation*}
r_i = \dfrac{\sum_{i=1}^m (x_i-\bar{x})(c-\bar{c})}{\sqrt{(\sum_{i=1}^m x_i-\bar{x})(\sum_{i=1}^m c-\bar{c})}}
\end{equation*}
where $m$ is the number of training examples, $x_i$ is the value of feature
$i$, and $\bar{x}$ and $\bar{c}$ are the arithmetic averages of the feature
values and the class values respectively.

\subsection{Information Gain}
We can also measure the IG of each feature as follows.
If we have a class $C$ and a feature $F$, the entropy of the
class before and after observing the feature is:
\begin{equation*}
\begin{aligned}
& \mathrm{entropy}(C) = - \sum_{c \in C} p(c) \mathrm{log}_2 p(c) \\
& \mathrm{entropy}(C|F) = - \sum_{f \in F} p(f) \sum_{c \in C} p(c|f) \mathrm{log}_2 p(c|f)
\end{aligned}
\end{equation*}
The decrease in entropy reflects the gain in information provided by the
feature and is given by:
\begin{equation*}
IG(C|F) = \mathrm{entropy}(C) - \mathrm{entropy}(C|F)
\end{equation*}

@@

\subsection{1R}
Another way to evaluate the relevance of features is to use the idea behind the
1R classifier described in Section. Recall that the 1R
classifier constructs simple 1-rules based on each feature and uses the feature
that produces the lowest error rate (that is, the highest accuracy) on the data
set to classify unseen examples. These accuracy rates can be calculated for
each feature, which are then used to rank the features from highest to lowest
accuracy.

We can also select a fixed number of features or all features above a certain
1R accuracy threshold. This threshold is a slightly different quantity to the
threshold of the correlation coefficient and information gain methods, as it
is an accuracy, expressed in percentages. Our first threshold is the Zero-R
accuracy for the trauma data set, which is the lower of the two Zero-R
accuracies on our two data sets. We increase this percentage by in increasing
increments to obtain a range of thresholds up to nearly 97\% accuracy (which
discards all features for both data sets). Using the same thresholds across
both data sets allows us to make valid comparisons of the effect of feature
selection on performance.

\subsection{Wrapper-based Feature Selection}
The feature selection methods described above (CFS, Pearson correlation and
1R) have been described as \textit{filter} approaches to attribute selection.
Kohavi and John proposed another approach, the \textit{wrapper} approach:
instead of performing feature selection independently of the learning
algorithm using some measure of ``relevance'',
the learning algorithm itself is used to evaluate the goodness of
subsets of features \cite{Kohavi1997}.

@@

\section{Evaluation Method}
Our evaluation was carried out using a trauma-specific data set
containing 2546 records collected from 2007--2011 collected at a major hospital
in Sydney, Australia.
\subsection{Classifiers}
We tested the performance of Zero-R, 1R, Na\"{i}ve Bayes, C4.5 decision trees,
support vector machines, multi-layer perceptrons, $k$-nearest neighbour with 1
and 20 nearest neighbours, K* and logistic regression, as well as our proposed
Ranked Distance Nearest Neighbour.

\subsection{Procedure}
The steps that we took to evaluate the performance of the various classifiers
and feature selection methods were:
\begin{enumerate}
\item For each data set, we evaluated the effect of the 4 automatic feature
selection methods on each of the 8 classifiers listed above, giving us 32
configurations for each data set, over
4 sets. Evaluation was performed using ten runs of \textit{ten-fold stratified
cross-validation}.
\item For the two trauma LOS data sets, we additionally evaluated the two
manual feature selection methods with all 8 classifiers. Recall from Section
\ref{sec:manual} that one of the feature sets is from the key previous work
on trauma LOS prediction that we use as a baseline, and the other feature set
is a set of features included through consultation with a domain expert.
\end{enumerate}

@@

>>>>

\documentclass{article}

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue,filecolor=blue,urlcolor=blue,pdfborder={0 0 0}]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{multicol}
\usepackage{natbib}
\usepackage{subcaption}

@@

\renewcommand{\abstractname}{Summary}

\begin{abstract}
Objective:

Materials and methods:

Results:

Conclusions:

Keywords:

\end{abstract}

@@

\section{Nearest Neighbour Classification}
In this section we will describe the $k$-nearest neighbour ($k$-NN) algorithm
and discuss some extensions which we use in our work. Additionally, we will
describe our contribution, which is an extension of the distance metric for
finding the nearest neighbours.

\subsection{Basic Algorithm}
The $k$-nearest neighbour classifier is an example of an
\textit{instance-based} learning algorithm. This is because it does not, unlike
the other classifiers described in this section, attempt to deduce or
generalise a relationship between the features and the class: it simply stores
all training examples and classifies an unseen example by finding the $k$
``nearest'' examples (or \textit{instances}) and assigning the majority class
to the new example. In the simplest form of the algorithm, $k$ is chosen to be
1 and the Euclidean distance is used to measure the closeness of neighbours:
\begin{equation*}
\mathrm{Distance} = \sqrt{(x_1-y_1)^2 + (x_2-y_2)^2 + \ldots + (x_n-y_n)^2}
\end{equation*}
where $x_i,y_i$ are the values of the $i$-th feature for two different feature
vectors (or instances).

Like the Na\"{i}ve Bayes classifier, $k$-NN allows all features to be taken
into account equally when classifying a new example. However, since feature
values often span different ranges and units of measurement, a particularly
large range of values for a feature would contribute much more to the distance
than another feature that spans a lower set of values.
To avoid this, it is important to \textit{normalise} the
values of each numeric feature before using $k$-NN: for each feature $i$, we
calculate the normalised value of the feature for each training example. This
is done using the following relationship:
\begin{equation*}
a_i = s\left(\dfrac{v_i - \mathrm{min }v_i}{\mathrm{max }v_i - \mathrm{min }v_i}\right) + t
\end{equation*}
where $a_i$ is the normalised feature value, $v_i$ is the actual value in the
data set, max $v_i$ and min $v_i$ are taken over all training examples, $s$ is
a scaling factor and $t$ indicates how much to translate the range by. If $s=1$
and $t=0$, then the range of normalised values will be in $[0,1]$.

\subsection{Existing and Proposed Extensions}
There have been many extensions proposed to the $k$-NN algorithm described
above. These usually seek to find a more suitable distance metric for the
problem domain at hand, or to reduce the storage requirements of the original
$k$-NN algorithm (since the classifier makes no attempt to generalise from the
examples, all of them must be stored).
We will focus on two algorithms that extend the standard Euclidean distance:
K* and our proposed approach, Ranked Distance Nearest Neighbour.

\subsubsection{K*: Instance-based Classifier with Entropic Distance Function}
Cleary and Trigg developed K*, an instance-based classifier that uses entropy
to measure the distances between instances \cite{Cleary1995}. They consider
the distance between two instances to be the complexity of transforming one
into another. Define a sequence of transformations on an instance:
\begin{equation*}
\bar{t}(a) = t_n(t_{n-1}(\ldots t_1(a)\ldots)), \bar{t} \in T
\end{equation*}
where $\bar{t} = t_1,\ldots,t_n$. We can define a probability function $p$
that gives the probabilities of these transformations occurring, and then we
can define a probability function that gives the probabilities of all paths
(transformations) from example $a$ to $b$:
\begin{equation*}
P^*(b|a) = \sum_{\bar{t}\in T:\bar{t}(a)=b} p(\bar{t})
\end{equation*}
Within this framework, we can define transformations for both real-valued and
categorical features, which allows K* to work with all types of features.
The K* function is then defined as:
\begin{equation*}
K^*(b|a) = -\mathrm{log}_2P^*(b|a)
\end{equation*}

To calculate the probability of an instance $a$ being in a class $C$, we sum
the probabilities from $a$ to each instance of $C$, and repeat for each class.
The predicted class will be the one with the highest probability:
\begin{equation*}
P^*(C|a) = \sum_{b\in C} P^*(b|a)
\end{equation*}

We use K* as a comparison to the basic NN algorithm and also as a comparison
to our proposed Ranked Distance Nearest Neighbour algorithm. Additionally,
K* has not been yet been applied in predicting LOS, and its performance in this
domain is worth investigating.

\subsubsection{Ranked Distance Nearest Neighbour}
\paragraph{Intuition}
As we mentioned earlier, all features are considered equally in a $k$-NN
classifier, which implies that all features are equally important in predicting
the class. In practice, however, this is not always the case: to predict the
LOS of a trauma patient, the severity of their injury should affect the final
LOS more than whether or not they can speak English (which could be just two
of many features that are recorded about a patient). Our contribution is a
new NN algorithm that uses a modified distance function which takes into
account the relative importance of all features called Ranked Distance
Nearest Neighbour (RD).

\paragraph{Mathematical formulation}
Given two instances $\mathbf{x} = (x_1,x_2,\ldots,x_n)$ and
$\mathbf{y} = (y_1,y_2,\ldots,y_n)$, the distance between them is given by:
\begin{equation*}
\mathrm{Distance} = \sum_{i=1}^n w_i |x_i-y_i|
\end{equation*}
where the $w_i$s weight the contribution of the $i$-th feature to the overall
distance that is computed by the $k$-NN algorithm. We assume that all feature
values have been normalised, as described above.

\paragraph{Assignment of weights}
The weights $w_i$ can be tuned to match the particular problem that is being
investigated. For our LOS classification problem, we will consider one way
of weighting features.
First, we compute the correlation coefficients as described in the point above.
This gives us a ranking of the importance of each feature to the prediction of
the class, with 1 being the highest rank (indicating the most importance) and
$n$ being the lowest rank (indicating the least importance).
Instead of using the values of the correlation directly to weight
the distance contribution of each feature, we specify a function:
\begin{equation*}
f : \mathrm{Rank} \rightarrow \mathbb{R}, \mathrm{Rank} \in \{1,2,\ldots,n\}
\end{equation*}
that describes
how the weights of the features vary with the rank. This function should
inituitively be non-increasing and should produce lower values when the rank
number is greater, meaning that less important features have a lower weight
in the distance calculation.
Note that choosing $f(\mathrm{Rank}) = k$ for any non-zero constant $k$
results in all features contributing equally to the distance calculation.
We will use $f(\mathrm{Rank}) = \frac{1}{\mathrm{Rank}^c}$, where
$c$ varies from 0 to 1.

\section{Evaluation}
\subsection{Data Set}
Our evaluation was carried out using a trauma-specific data set
containing 2546 records collected from 2007--2011 collected at a major hospital
in Sydney, Australia.

\subsection{Feature Selection}
Here we introduce the feature selection algorithms we evaluate in our work.
Several of these have not been investigated in previous work on length of
stay prediction.

\subsubsection{Pearson Correlation Coefficient}
The Pearson correlation coefficient measures the degree of linear correlation
between two variables. It is always in the range $[-1,1]$, where -1 is total
negative correlation, 0 is no (linear) correlation, and 1 is total positive
correlation. We compute the Pearson correlation between each feature and the
class using:
\begin{equation*}
r_i = \dfrac{\sum_{i=1}^m (x_i-\bar{x})(c-\bar{c})}{\sqrt{(\sum_{i=1}^m x_i-\bar{x})(\sum_{i=1}^m c-\bar{c})}}
\end{equation*}
where $m$ is the number of training examples, $x_i$ is the value of feature
$i$, and $\bar{x}$ and $\bar{c}$ are the arithmetic averages of the feature
values and the class values respectively.

\subsubsection{Information Gain}
We can also measure the IG of each feature as follows.
If we have a class $C$ and a feature $F$, the entropy of the
class before and after observing the feature is:
\begin{equation*}
\begin{aligned}
& \mathrm{entropy}(C) = - \sum_{c \in C} p(c) \mathrm{log}_2 p(c) \\
& \mathrm{entropy}(C|F) = - \sum_{f \in F} p(f) \sum_{c \in C} p(c|f) \mathrm{log}_2 p(c|f)
\end{aligned}
\end{equation*}
The decrease in entropy reflects the gain in information provided by the
feature and is given by:
\begin{equation*}
IG(C|F) = \mathrm{entropy}(C) - \mathrm{entropy}(C|F)
\end{equation*}

@@

\subsubsection{Wrapper-based Feature Selection}
The feature selection methods described above (CFS, Pearson correlation and
1R) have been described as \textit{filter} approaches to attribute selection.
Kohavi and John proposed another approach, the \textit{wrapper} approach:
instead of performing feature selection independently of the learning
algorithm using some measure of ``relevance'',
the learning algorithm itself is used to evaluate the goodness of
subsets of features \cite{Kohavi1997}.

@@

\subsection{Classification Algorithms}
We tested the performance of Zero-R, 1R, Na\"{i}ve Bayes, C4.5 decision trees,
support vector machines, multi-layer perceptrons, $k$-nearest neighbour with 1
and 20 nearest neighbours, K* and logistic regression, as well as our proposed
Ranked Distance Nearest Neighbour.

\subsection{Experimental Setup and Procedure}
The steps that we took to evaluate the performance of the various classifiers
and feature selection methods were:
\begin{enumerate}
\item For each data set, we evaluated the effect of the 4 automatic feature
selection methods on each of the 8 classifiers listed above, giving us 32
configurations for each data set, over
4 sets. Evaluation was performed using ten runs of \textit{ten-fold stratified
cross-validation}.
\item For the two trauma LOS data sets, we additionally evaluated the two
manual feature selection methods with all 8 classifiers. Recall from Section
\ref{sec:manual} that one of the feature sets is from the key previous work
on trauma LOS prediction that we use as a baseline, and the other feature set
is a set of features included through consultation with a domain expert.
\end{enumerate}

@@

\section*{Acknowledgements}
